Performance Evaluation of Big Data Processing Strategies for Neuroimaging

### 2018-11-26 paper.tex

\usepackage{graphicx}
\usepackage{color}
\usepackage[bookmarks=false]{hyperref}

\newcommand{\todo}[1]{\marginpar{\parbox{18mm}{\flushleft\tiny\color{red}\textbf{TODO}:
      #1}}}

@@

Big Data processing engines leverage data locality, 
in-memory computing, and lazy evaluation to reduce the impact of data 
manipulation on the performance of applications. From early MapReduce 
implementations~\cite{dean2008mapreduce} to Apache Spark~\cite{zaharia2016apache} and now 
Dask~\cite{rocklin2015dask}, these frameworks have become 
mainstream tools for data analytics. Meanwhile, several scientific 
domains including bioinformatics, physics or astronomy, have entered 
the Big Data era due to increasing data volumes and variety. 
However, the adoption of Big Data engines for scientific data analysis 
remains limited, perhaps due to the widespread availability of 
scientific processing engines such as Pegasus~\cite{deelman2005pegasus}, 
Taverna~\cite{oinn2004taverna}, and many others.

Scientific applications slightly differ from the typical Big Data use 
cases, which might explain the remaining gap between Big Data and 
scientific engines. While Big Data applications mostly target text 
processing (Web search, frequent pattern mining, recommender systems, 
etc~\cite{leskovec2014mining}), implemented in consistent software libraries, and 
running on clouds or dedicated commodity clusters with locality-aware 
file systems such as HDFS~\cite{shvachko2010hadoop}, scientific applications often 
involve binary data such as images and signals, processed by 
command-line tools using a mix of programming languages (C, Fortran, 
Python, shell scripts) and deployed on large, shared clusters where 
data is fetched from data nodes to compute nodes using shared file 
systems such as Lustre~\cite{schwan2003lustre}. Such differences in applications 
and infrastructures have important consequences. To mention only one, 
in-memory computing requires instrumentation to be applied to 
command-line tools.

Technological advances of the past decade, in particular page caching 
in the Linux kernel~\cite{love2010linux}, in-memory file systems 
(\texttt{tmpfs}) and memory-mapped files\todo{check that} might also 
explain the lack of adoption of Big Data engines for scientific 
applications.

In-memory computing would then be a feature provided by 
the operating system rather than by the engine itself. The frontier 
between these two components is blurred and needs to be clarified.

@@

Our study focuses on performance. We intentionally do not intend to 
compare Big Data and scientific data processing engines on the grounds 
of workflow language expressivity, provenance capture and 
representation, portability or reproducibility, which are otherwise 
critical concerns. Besides, our study of performance focuses on the 
impact of data writes and transfers. It purposely leaves out task 
scheduling to computing resources, to focus on the understanding of 
data writes and movement. Task scheduling will be part of our 
discussion though.

@@

@@

% pipeline description: map nodes and their limitations
% general description relevant to the experiment
% two plugins
% provenance
% file system caching
% memory management (bad?)

\subsection{Performance Models}

\subsubsection{In-Memory File System}

@@

% 15 cores per executor can lead to bad I/O throughput
% http://blog.cloudera.com/blog/2015/03/how-to-tune-your-apache-spark-jobs-part-2/

@@

The Linux kernel uses in-memory page caching to speed-up disk writes. 
When a process calls the \texttt{write()} system call, data is copied 
to a memory cache that is asynchronously written to disk by flusher 
threads when memory shrinks, ``dirty" (unwritten) data grows, or a 
process invokes the \texttt{sync()} system call~\cite{love2010linux}. 
The flushing of page cache to disk is also called the writeback 
process.

Page caching has a dramatic effect on I/O, reducing the duration of 
disk writes by several orders of magnitude. It is essentially a way to 
emulate in-memory computing at the kernel level, without requiring a 
dedicated engine. The size of the page cache, however, becomes a 
limitation when processes write faster than the disk bandwidth. When 
this happens, the page cache rapidly fills up and writes are limited by 
the disk write bandwidth -- typically 500~MB/s for random writes to 
contemporary SSDs  -- as if no page cache was involved.

@@

In general, an application should ensure that $\dot d$ remains negative 
or null, leading to the following inequality:
\begin{equation}
\frac{D}{C} \leq \frac{\delta}{\gamma} \label{eq:page-cache-inequality}
\end{equation}
This defines a data-compute ratio beyond which the page cache becomes 
asymptotically useless, which we will use to define our benchmarks. It 
should be noted that leveraging the page cache, and therefore ensuring 
that Equation~\ref{eq:page-cache-inequality} holds, has important 
performance implications: with page caching, the application will 
experience an almost negligible write time, while without page caching 
the write throughput will be that of the disk.

% Cache eviction: LRU/n.
% page caching is mentioned in https://ntrs.nasa.gov/archive/nasa/casi.ntrs.nasa.gov/20130001600.pdf too
% Note: writeback is dangerous as a system failure could result in data loss.

@@

@@

% Say that we don't use HDFS because we focus on clusters and HDFS on multi-tenant clusters is difficult (refer to hadoop on demand).

@@

@@

% fault-tolerance: do we have anything to discuss here?

>>>>

\usepackage{graphicx}
\usepackage{color}
\usepackage{colortbl} % for \rowcolor

\usepackage[bookmarks=false]{hyperref}

\newcommand{\todo}[1]{\marginpar{\parbox{18mm}{\flushleft\tiny\color{red}\textbf{TODO}:
      #1}}}
\definecolor{headcolor}{gray}{0.9}

@@

Big Data processing engines leverage data locality, 
in-memory computing, and lazy evaluation to reduce the impact of data 
manipulation on the performance of applications. \todo{We need to 
define data locality, in-memory computing and lazy evaluation somewhere 
in the intro.} From early MapReduce 
implementations~\cite{dean2008mapreduce} to Apache 
Spark~\cite{zaharia2016apache} and now Dask~\cite{rocklin2015dask}, 
these frameworks have become 
mainstream tools for data analytics. Meanwhile, several scientific 
domains including bioinformatics, physics or astronomy, have entered 
the Big Data era due to increasing data volumes and variety. 
However, the adoption of Big Data engines for scientific data analysis 
remains limited, perhaps due to the widespread availability of 
scientific processing engines such as Pegasus~\cite{deelman2005pegasus}, 
Taverna~\cite{oinn2004taverna}, and many others.

Scientific applications slightly differ from the typical Big Data use 
cases, which might explain the remaining gap between Big Data and 
scientific engines. While Big Data applications mostly target text 
processing (Web search, frequent pattern mining, recommender systems, 
etc~\cite{leskovec2014mining}), implemented in consistent software libraries, and 
running on clouds or dedicated commodity clusters with locality-aware 
file systems such as HDFS~\cite{shvachko2010hadoop}, scientific applications often 
involve binary data such as images and signals, processed by 
command-line tools using a mix of programming languages (C, Fortran, 
Python, shell scripts) and deployed on large, shared clusters where 
data is fetched from data nodes to compute nodes using shared file 
systems such as Lustre~\cite{schwan2003lustre}. Such differences in 
applications and infrastructures have important consequences. To 
mention only one, in-memory computing requires instrumentation to be 
applied to command-line tools. \todo{Also mention that performance is 
only one of the concerns of scientific processing engines, which might
explain why Big Data engines are overlooked.}

Technological advances of the past decade, in particular page caching 
in the Linux kernel~\cite{love2010linux}, in-memory file systems 
(\texttt{tmpfs}) and memory-mapped files\todo{check that} might also 
explain the lack of adoption of Big Data engines for scientific 
applications. 
In-memory computing would then be a feature provided by 
the operating system rather than by the engine itself. The frontier 
between these two components is blurred and needs to be clarified.

@@

Our study focuses on performance. We intentionally do not intend to 
compare Big Data and scientific data processing engines on the grounds 
of workflow language expressivity, fault-tolerance, provenance capture and 
representation, portability or reproducibility, which are otherwise 
critical concerns. Besides, our study of performance focuses on the 
impact of data writes and transfers. It purposely leaves out task 
scheduling to computing resources, to focus on the understanding of 
data writes and movement. Task scheduling will be part of our 
discussion though.

@@

\newpage % Remove this when abstract is written

@@

% pipeline description: map nodes and their limitations
% general description relevant to the experiment
% two plugins
% provenance
% file system caching
% memory management (bad?)
% Nipype only has data locality becausae we use multiproc. Data locality is emulated by manual decisions on scheduling.

\subsection{Data Storage Locations}

Data storage location is critical to the performance of Big Data 
applications on HPC clusters. Data may reside in the engine memory, on 
a file system whose content reside in virtual memory (for instance 
\texttt{tmpfs}), on disks local to the processing node, or on a shared 
file system. Table~\ref{table:features} summarizes the Big Data 
strategies that can be used depending on the data location. The 
remainder of this Section explains this Table and provides related 
performance models.
\begin{table}
\centering
\begin{tabular}{c|ccc}
   \rowcolor{headcolor}
    Data Location                 & In-Memory     & Data Locality & Lazy \\
    \rowcolor{headcolor}
                                  & Computing     &               & Evaluation      \\
                                  \hline          
In-memory                         &   Y           & Y             & Spark               \\
\texttt{tmpfs}                    &   Y           & Y             & Spark           \\
Local Disk                        & Page Caching  & Y             & Spark           \\
Shared File System                & Page Caching  & N            & Spark          
% data locality is somehow present in shared fs due to page caching again.
\end{tabular}
\caption{Big Data strategies on a shared HPC cluster.}
\label{table:features}
\end{table}

\subsubsection{In-Engine-Memory and In-Memory File System}

@@

@@

Storing data on local disks obviously enables data locality as tasks 
executed on the node where their input files reside do not require to 
transfer input or output data. However, in absence of a more specific 
file system such as the Hadoop Distributed File System (HDFS) to handle 
file replication across computing nodes, data locality comes at the price
of stringent scheduling restrictions, as tasks can only be scheduled to the
single node that contains their input data.
% this brings data locality but not at the level of HDFS because replication isn't there and 
% there is only 1 single option to schedule a task while preserving data locality.

The performance of local disk accesses is strongly dependent on the 
page caching mechanism provided by the Linux kernel, described in great 
details in~\cite{love2010linux}. To summarize, data read from disk 
remains cached in memory until evicted by an LRU (Least Recently Used) 
strategy. When a process invokes the \texttt{read()} system call, the 
kernel will return the data directly from memory if the requested data 
lies in the page cache. Such cache hits drastically speed-up data 
reads, by masking the disk latency and limited bandwidth behind a 
memory buffer. In effect, page caching provides in-memory computing 
transparently to the processing engine. However, page cache eviction 
strategies currently cannot be controlled by the application 
\todo{check that}, which prevents processing engines to anticipate 
reads by preloading the cache. Scheduling strategies might be designed 
to maximize cache hits though. For instance, we hypothesize that lazy 
evaluation would result in more cache hits by scheduling data-dependent 
tasks on the same node.

Page caching has a more dramatic effect on disk writes, reducing their 
duration by several orders of magnitude. When a process calls the 
\texttt{write()} system call, data is copied to a memory cache that is 
asynchronously written to disk by flusher threads when memory shrinks, 
``dirty" (unwritten) data grows, or a 
process invokes the \texttt{sync()} system call. 
This asynchronous flushing of the page cache is called 
\emph{writeback}.
%~ It should be noted that writeback may lead to data 
%~ loss in case of system failure.

Page caching is essentially a way to 
emulate in-memory computing at the kernel level, without requiring a 
dedicated engine. The size of the page cache, however, becomes a 
limitation when processes write faster than the disk bandwidth. When 
this happens, the page cache rapidly fills up and writes are limited by 
the disk write bandwidth -- about 500~MB/s for random writes to 
contemporary SSDs  -- as if no page cache was involved.

@@

In general, an application should ensure that $\dot d$ remains negative 
or null, leading to the following inequality:
\begin{equation}
\frac{D}{C} \leq \frac{\delta}{\gamma} \label{eq:page-cache-inequality}
\end{equation}
This defines a data-compute ratio beyond which the page cache becomes 
asymptotically useless. We will use this ratio to define our benchmarks. It 
should be noted that leveraging the page cache, and therefore ensuring 
that Equation~\ref{eq:page-cache-inequality} holds, has important 
performance implications: with page caching, the application will 
experience negligible write times, while without page caching 
the write throughput will be that of the disk.

% Cache eviction: LRU/n.

\todo{How about reads?}

@@

We model a shared file system using its global apparent bandwidth 
$\Delta$, shared by all concurrent processes in the system. We are 
aware that such a simplistic model does not describe at all the 
intricacies of systems such as Lustre. In particular, metadata 
management, RPC protocol optimizations and storage optimizations are 
all covered under the apparent bandwidth. We do, however, consider the 
effect of page caching in shared file systems too, since in Linux 
writes to network-mounted volumes benefit from this feature too.

As in the local disk model, we note that page caching will only be 
useful when the flush bandwidth is greater than the write throughput of 
the application, that is:
\begin{equation}
\frac{D}{C} \leq \frac{\Delta}{\Gamma}, \label{eq:page-cache-sharedfs}
\end{equation}
where $\Gamma$ is the max number of concurrent jobs in the cluster. 
Note that $\frac{\Delta}{\Gamma}$ will be much lower than 
$\frac{\delta}{\gamma}$ in Equation~\ref{eq:page-cache-inequality}.

@@

% Say that we don't use HDFS because we focus on multi-tenant clusters and HDFS on multi-tenant clusters is difficult (refer to hadoop on demand).

@@

A common recommendation in Spark is to limit the number of cores per 
executors to 5, to preserve a good I/O 
throughput\footnote{\url{http://blog.cloudera.com/blog/2015/03/how-to-tune-your-apache-spark-jobs-part-2}}. 
We believe that throughput degradation observed with more than 5 cores 
per executor might be coming from full page caches.

@@

% fault-tolerance: if you loose a node, you loose the data when it's written on disk or in memory.
%                  then the only solution is to recompute, which spark would do automatically.
%                  when nipype is run on lustre, it would avoid recomputing existing files, which Spark doesn't do.
%                   this niype feature is lost when not using lustre.

% Comment on the practical implications of using tmpfs or local disk: we might not get all the intermediary files.

% mention burst buffers or heterogeneous storage managers as potential solutions to write to local disk, 
% i.e., to benefit from data locality while getting your results on Lustre.
